Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
pagebreak after abstract
Some fiddling.

\maketitle

@@

\item \textbf{Marlon Pierce: Patching It Up, Pulling It
Forward~\cite{Pierce_poster}.} Pierce discussed how open open source is. Open
software needs a diverse, openly governed community behind it, just as it needs
open licensing and a public code repository. To probe the level of governance
within open source projects, he and his co-authors (S.~Marru and C.~Mattmann)
suggested a contest to encourage individual developers to submit
patches and requests to projects that are important to them. This simple
mechanism shall expose several governance mechanisms, such as how easy it is for
independent developers to communicate with project leadership, how projects
accept and license third-party contributions, and how projects make
decisions such as granting source tree write access.

@@

\item \textbf{James Howison: Retract all Bit-Rotten Publications. }%~\cite{Howison_poster}
Howison sought to provoke discussion by proposing that papers whose workflows
are not kept current with the changing software ecosystem should be
automatically retracted. This would create an incentive for authors to keep
their software current and usable, rather than the current situation in which
every potential user has to do this individually. A softer version of the
proposal would identify papers whose software workflow has become bit-rotten and
allow others to keep the code up to date, either adding them as new authors of
the paper or providing credit in some other form for their academic service.

@@

The first group discussed ways in which reproducibility of papers could be
improved. A consensus surrounded the provision of examples: demonstrating to
others how to achieve reproducibility. Major public funding investments go into
research that heavily relies on software to be reproducible, hence lack thereof
is raising major concerns.

@@

Discovery of relevant software is still fundamental issue: we need standard
vocabularies and metadata, better tools, and approaches that are sustainable.
Publications are an easy entry point for locating suitable software, but should
the publishers lead the way, or is this the responsibility of ``the community''?
Some communities have had significant initiatives to improve software discovery,
e.g., the NIH Software Discovery Index\footnote{NIH Software Discovery Index:
\url{http://softwarediscoveryindex.org/}}. Likewise, there were examples of
journals which had made software more discoverable: ACM Transactions on Software
offers a reproducibility review; the Journal of Biostatistics has an opt-in to
provide code and a certification mark if it can be run; the Journal of Open
Research Software requires software to be deposited in suitable repositories and
referenced.

An issue around the sustainability of software catalogues is that their
usefulness often depends on the domain. For instance, in the bio-sciences, there
are more homogeneous data and standard shred code. In areas like ecology,
codes are very specific to the problem, meaning that the level of re-use might
be at a very general statistical level of abstraction, but then every research
use is highly customized.

@@

This group aimed to come up with a categorization of journals which published
software papers. Starting from the list of
journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}}
maintained by the Software Sustainability Institute, the group chose seven
journals and looked at their advice to authors and reviewers. These were the
Journal of Open Research
Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}},
PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}},
Journal of Statistical
Software\footnote{\url{http://www.jstatsoft.org/instructions}}. Methods in
Ecology and
Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}},
Transactions of Mathematical
Software\footnote{\url{http://toms.acm.org/Authors.html}},
GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS
Computational
Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.

From these a set of common categories were synthesized, against which all
journals could be compared:

@@

The two non-self-taught paths were direct learning and indirect through other
software (learning by example). The group recognized that teaching about testing
needs to start early in projects and careers. We need to teach people how to
test short bits of code, rather than waiting until they have thousands of lines
of code and being surprised when they don't write tests. The group generated a
set of practical suggestions for improving adoption of code testing.

Established software projects can encourage the latter through a simple rule of
accepting only software patches with associated tests and through good
documentation of testing practices and requirements.

>>>>

\maketitle
\newpage

@@

\item \textbf{Marlon Pierce: Patching It Up, Pulling It
Forward~\cite{Pierce_poster}.} Pierce discussed how open open source is. Open
software needs a diverse, openly governed community behind it, just as it needs
open licensing and a public code repository. To probe the level of governance
within open source projects, he and his collaborators %(S.~Marru and C.~Mattmann)
suggested a contest to encourage individual developers to submit
patches and requests to projects that are important to them. This simple
mechanism shall expose several governance mechanisms, such as how easy it is for
independent developers to communicate with project leadership, how projects
accept and license third-party contributions, and how projects make
decisions such as granting source tree write access.

@@

\item \textbf{James Howison: Retract all Bit-Rotten Publications. }%~\cite{Howison_poster}
Howison sought to provoke discussion by proposing that papers whose workflows
are not kept current with the changing software ecosystem should be
automatically retracted. This would create an incentive for authors to keep
their software current and usable, rather than the current situation in which
every potential user has to do this individually. A softer version of the
proposal would identify papers whose software workflow has become bit-rotten and
allow others to keep the code up to date, either adding them as new authors of
the paper or providing credit in some other form. %for their academic service.

@@

The first group discussed ways in which reproducibility of papers could be
improved. A consensus surrounded the provision of examples: demonstrating to
others how to achieve reproducibility. Major public funding investments go into
research that heavily relies on reproducible software, hence lack thereof
is raising major concerns.

@@

Discovery of relevant software is still a fundamental issue: we need standard
vocabularies and metadata, better tools, and approaches that are sustainable.
Publications are an easy entry point for locating suitable software, but should
the publishers lead the way, or is this the responsibility of ``the community''?
Some communities have had significant initiatives to improve software discovery,
e.g., the NIH Software Discovery Index\footnote{NIH Software Discovery Index:
\url{http://softwarediscoveryindex.org/}}. Likewise, there were examples of
journals which had made software more discoverable: ACM Transactions on Software
offers a reproducibility review; the Journal of Biostatistics has an opt-in to
provide code and a certification mark if it can be run; the Journal of Open
Research Software requires software to be deposited in suitable repositories and
referenced.

An issue around the sustainability of software catalogues is that their
usefulness often depends on the domain. For instance, in the bio-sciences, there
are more homogeneous data and standard shred code. In areas like ecology,
code is often specific to a problem, meaning that the level of re-use might
be at a general statistical level of abstraction, but then every research
use is highly customized.

@@

This group aimed to come up with a categorization of journals which published
software papers. Starting from the list of
journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}}
maintained by the Software Sustainability Institute, the group chose seven
journals and looked at their advice to authors and reviewers. These were the
Journal of Open Research
Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}},
PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}},
Journal of Statistical
Software\footnote{\url{http://www.jstatsoft.org/instructions}}. Methods in
Ecology and
Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}},
Transactions of Mathematical
Software\footnote{\url{http://toms.acm.org/Authors.html}},
GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS
Computational
Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.
%
From these a set of common categories were synthesized, against which all
journals could be compared:

@@

The two non-self-taught paths were direct learning and indirect through other
software (learning by example). The group recognized that teaching about testing
needs to start early in projects and careers. We need to teach people how to
test short bits of code, rather than waiting until they have thousands of lines
of code and being surprised when they don't write tests. The group generated a
set of practical suggestions for improving adoption of code testing.
%
Established software projects can encourage the latter through a simple rule of
accepting only software patches with associated tests and through good
documentation of testing practices and requirements.
